---
title: "Figure S5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script reproduces the analyses for Figure S5.

```{r}
library(BASiCS)
library(ggplot2)
library(openxlsx)
library(pheatmap)
```

Load in the chains.

```{r}
MCMC.CAST.naive <- readRDS("../../../MCMCs/chain_CAST.naive.Rds")
MCMC.CAST.active <- readRDS("../../../MCMCs/chain_CAST.active.Rds")

# Read in metadata file
CD4.metadata <- read.table("metadata_file.txt", header = TRUE, sep = "\t")
rownames(CD4.metadata) <- CD4.metadata$X

norm.counts <- readRDS("../../../Results/norm_counts.rds")
CD4.metadata <- CD4.metadata[colnames(norm.counts),]

CD4.raw <- read.table("raw_data.txt", header = TRUE, sep = "\t")
```

Calculate tSNE on normalized counts

```{r}
naive_cells <- norm.counts[,CD4.metadata$Stimulus == "Unstimulated" &
                          (CD4.metadata$Individuals == "CAST young 1" |
                             CD4.metadata$Individuals == "CAST young 2")]
colnames(naive_cells) <- paste("Naive", colnames(naive_cells), sep = "_") 

active_cells <- norm.counts[,CD4.metadata$Stimulus == "Active" &
                          (CD4.metadata$Individuals == "CAST young 1" |
                             CD4.metadata$Individuals == "CAST young 2")]
colnames(active_cells) <- paste("Active", colnames(active_cells), sep = "_") 


cur_cells <- cbind(naive_cells, active_cells)

# Compute tSNE
library(Rtsne)
set.seed(1234)
tsne <- Rtsne(t(log10(cur_cells + 1)), perplexity = 10)

tsne.df <- data.frame(tsne.1 = tsne$Y[,1], tsne.2 = tsne$Y[,2], 
                      activation = sapply(colnames(cur_cells), 
                                          function(n){unlist(strsplit(n, split = "_"))[1]}))

# Plot tSNE
naive_active_tsne <- ggplot(data = tsne.df, aes(tsne.1, tsne.2)) + 
  geom_point(aes(shape = activation), size = 4) +
  scale_shape_manual(values = c("|", "-")) + 
  theme_minimal() +
  xlab("tSNE 1") +
  ylab("tSNE 2") 
ggsave("../../Results/EFDR5/Fig_S5/A.pdf", naive_active_tsne, width = 5, height = 4)
```

Differential mean expression testing.

```{r}
OffSetCorrection <- function(MCMC1, MCMC2){
  median(rowSums(MCMC1@parameters$mu)/rowSums(MCMC2@parameters$mu)) 
}
OffSetCorrection1 <- function(MCMC1, MCMC2){
  median(rowMedians(MCMC1@parameters$mu)/rowMedians(MCMC2@parameters$mu))
}
# To properly exclude genes, we need to account or the offset
Offset <- OffSetCorrection(MCMC.CAST.naive, MCMC.CAST.active); Offset
Offset1 <- OffSetCorrection1(MCMC.CAST.naive, MCMC.CAST.active); Offset1

par(mfrow = c(1,3))
boxplot(cbind(colMedians(MCMC.CAST.naive@parameters$mu),
              colMedians(MCMC.CAST.active@parameters$mu)), 
        log = "y", names = c("naive", "active"), frame = FALSE,
        ylab = "Mean expression estimate", main = "Before offset (mean-based)")
boxplot(cbind(colMedians(MCMC.CAST.naive@parameters$mu) / Offset,
              colMedians(MCMC.CAST.active@parameters$mu)), 
        log = "y", names = c("naive", "active"), frame = FALSE,
        ylab = "Mean expression estimate", main = "After offset (mean-based)")
boxplot(cbind(colMedians(MCMC.CAST.naive@parameters$mu) / Offset1,
              colMedians(MCMC.CAST.active@parameters$mu)), 
        log = "y", names = c("naive", "active"), frame = FALSE,
        ylab = "Mean expression estimate", main = "After offset (median-based)")
```

For these data, we observe that a handful of genes with high expression in 
active cells distored the offset calculation as it is based on the average of 
mean expression parameters per cell group. As a more robust alternative, we 
replace the offset calculation to use medians instead of sums. This change does 
not alter the results for the other comparisons shown in this analysis.

```{r}
# Manual application of offset correction
MCMC.CAST.naive1 <- MCMC.CAST.naive
MCMC.CAST.naive1@parameters$mu <- MCMC.CAST.naive1@parameters$mu/Offset1
MCMC.CAST.naive1@parameters$phi <- MCMC.CAST.naive1@parameters$phi * Offset1

genes_select <- (colMedians(MCMC.CAST.naive@parameters$mu) > 1*Offset1 | 
                   colMedians(MCMC.CAST.active@parameters$mu) > 1)

Test_DE <- BASiCS_TestDE(Chain1 = MCMC.CAST.naive1, 
                              Chain2 = MCMC.CAST.active,
                              EpsilonM = 2,
                              GroupLabel1 = "Naive", 
                              GroupLabel2 = "Active", 
                              Plot = FALSE,
                              PlotOffset = FALSE, EFDR_M = 0.05,
                              Offset = FALSE,
                              GenesSelect = genes_select)

Test_DE.LFC0 <- BASiCS_TestDE(Chain1 = MCMC.CAST.naive1, 
                              Chain2 = MCMC.CAST.active,
                              EpsilonM = 0,
                              GroupLabel1 = "Naive", 
                              GroupLabel2 = "Active", 
                              Plot = FALSE,
                              PlotOffset = FALSE, EFDR_M = 0.05,
                              Offset = FALSE,
                              GenesSelect = genes_select)

# For visualization
cur_df <- data.frame(naive = Test_DE$TableMean$Mean1,
                     active = Test_DE$TableMean$Mean2,
                     testing = ifelse(Test_DE$TableMean$ResultDiffMean == "Naive+", "Naive",
                                      ifelse(Test_DE$TableMean$ResultDiffMean == "Active+", "Active", 
                                             ifelse(Test_DE.LFC0$TableMean$ResultDiffMean == "NoDiff", "For_variability", "NoDiff"))))
Active_naive.plot <- ggplot(cur_df) +
  geom_point(aes(log10(naive + 1), log10(active + 1), colour = testing)) + 
  scale_colour_manual(values = c("Naive" = "dark blue",
                                 "Active" = "dark red",
                                 "NoDiff" = "grey",
                                 "For_variability" = "black")) + theme_minimal()
ggsave("../../Results/EFDR5/Fig_S5/B.pdf", width = 7, height = 5)

# variability
ind_nochange <- Test_DE.LFC0$TableMean$ResultDiffMean == "NoDiff"

boxplot_variability <- ggplot(data.frame(Condition = factor(c(rep("Naive", sum(ind_nochange)),
                                rep("Active", sum(ind_nochange))),
                                levels = c("Naive", "Active")),
                  values = c(log10(Test_DE.LFC0$TableDisp$Disp1[ind_nochange]),
                             log10(Test_DE.LFC0$TableDisp$Disp2[ind_nochange])))) +
  geom_boxplot(aes(Condition, values), outlier.shape = NA) + ylim(c(0,2.5)) + theme_minimal()
ggsave("../../Results/EFDR5/Fig_S5/C.pdf", boxplot_variability, width = 2.5, height = 3)

wilcox.test(log10(Test_DE.LFC0$TableDisp$Disp1[ind_nochange]),
            log10(Test_DE.LFC0$TableDisp$Disp2[ind_nochange]))
```

Fraction of cells in which naive and active specific genes are expressed.

First for the naive.

```{r}
# Naive genes
cur_naive <- CD4.raw[,CD4.metadata$Strain == "Mus musculus castaneus" & 
                       CD4.metadata$Age == "Young" &
                       CD4.metadata$Stimulus == "Unstimulated" &
                       (CD4.metadata$Individuals == "CAST young 1" | 
                          CD4.metadata$Individuals == "CAST young 2")]
naive_genes <- Test_DE$TableMean$GeneName[Test_DE$TableMean$ResultDiffMean == "Naive+"]
cell_fraction.naive <- apply(cur_naive[naive_genes,], 1, function(n){sum(n > 0)/ncol(cur_naive)})
median(cell_fraction.naive)
# 0.1714

set.seed(12345)
pdf("../../Results/EFDR5/Fig_S5/Di.pdf", width = 7, height = 4)
hist(apply(cur_naive[naive_genes[sample(1:length(naive_genes), 400)],], 1, function(n){length(which(n > 0))/ncol(cur_naive)}), breaks = 30, xlim = c(0,1),ylim = c(0,100))
abline(v = median(cell_fraction.naive))
dev.off()
```

Now for the active

```{r}
# Active genes
cur_active <- CD4.raw[,CD4.metadata$Strain == "Mus musculus castaneus" & 
                       CD4.metadata$Age == "Young" &
                       CD4.metadata$Stimulus == "Active" &
                       (CD4.metadata$Individuals == "CAST young 1" | 
                          CD4.metadata$Individuals == "CAST young 2")]
active_genes <- Test_DE$TableMean$GeneName[Test_DE$TableMean$ResultDiffMean == "Active+"]
cell_fraction.active <- apply(cur_active[active_genes,], 1, function(n){sum(n > 0)/ncol(cur_active)})
median(cell_fraction.active)
# 0.314

set.seed(12345)
pdf("../../Results/EFDR5/Fig_S5/Dii.pdf", width = 7, height = 4)
hist(apply(cur_active[active_genes[sample(1:length(active_genes), 400)],], 1, function(n){length(which(n > 0))/ncol(cur_active)}), breaks = 30, xlim = c(0,1),ylim = c(0,100))
abline(v = median(cell_fraction.active))
dev.off()

wilcox.test(cell_fraction.naive, cell_fraction.active)
```




